Enforced Invariant Confluence

### 2018-02-27 doc/paper_redux/interactive_decision_procedure.tex
Finished interactive decision procedure section.

\newcommand{\comment}[1]{\State \textcolor{flatdenim}{\texttt{//} #1}}
\begin{algorithm}
  \caption{Interactive \Iconfluence{} Decision Procedure}%
  \algolabel{InteractiveDecisionProcedure}
  \begin{algorithmic}
    \Function{IsIconfluent}{$O$, $T$, $I$}
      \State \Return \Call{Helper}{$O$, $T$, $I$, $\emptyset$, $\emptyset$}
    \EndFunction

@@

    \comment{$R$ is a set of reachable states.}
    \comment{$NR$ is a set of non-reachable states.}
    \Function{Helper}{$O$, $T$, $I$, $R$, $NR$}
      \State closed, $s_1$, $s_2$ $\gets$
               \Call{IsIclosed}{$O$, $T$, $(I \cap R) - NR$}
      \If {closed $=$ unknown}
        \State \Return unknown
      \ElsIf {closed $=$ yes}
        \State \Return{yes}
      \Else
        \State Augment $R$ and $NR$ with random search and user input
        \State Make sure $s_1$ and $s_2$ are included in $R$ or $NR$
        \If{$s_1, s_2 \in R$}
          \State \Return no
        \Else
          \State \Return \Call{Helper}{$O$, $T$, $I$, $R$, $NR$}
        \EndIf
      \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

This example motivates \algoref{InteractiveDecisionProcedure}, a general
purpose interactive \Iconfluence{} decision procedure in which a user
iteratively refines the invariants until the decision procedure terminates.
Given an object $O$, a set of transactions $T$, and an invariant $I$, we call
\textsc{IsIconfluent}$(O, T, I)$ to determine if $T$ is \Iconfluent{}.
\textsc{IsIconfluent} assumes access to an \Iclosed{} decision procedure
\textsc{IsIclosed}$(O, T, I)$ which returns whether $O$ is \Iconfluent{} (yes,
no, or unknown). If it returns no, it also returns a counterexample $s_1$ and
$s_2$.

\todo{%
  Describe \textsc{IsIconfluent}. Describe how we can implement
  \textsc{IsIclosed} with Z3 and how users can specify $R$ and $NR$ with
  invariants.
}

>>>>

\newcommand{\comment}[1]{\State \textcolor{flatdenim}{\texttt{//} #1}}
\begin{algorithm}
  \caption{Interactive \Iconfluence{} Decision Procedure}%
  \algolabel{InteractiveDecisionProcedure}
  \begin{algorithmic}
    \Function{IsIconfluent}{$O$, $T$, $I$}
      \State \Return \Call{Helper}{$O$, $T$, $I$, $\set{s_0}$, $\emptyset$}
    \EndFunction

@@

    \comment{$R$ is a set of reachable states.}
    \comment{$NR$ is a set of non-reachable states.}
    \Function{Helper}{$O$, $T$, $I$, $R$, $NR$}
      \State closed, $s_1$, $s_2$ $\gets$
               \Call{IsIclosed}{$O$, $T$, $I - NR$}
      \If {closed $=$ unknown}
        \State \Return unknown
      \ElsIf {closed $=$ yes}
        \State \Return{yes}
      \Else
        \State Augment $R$ and $NR$ with random search and user input
        \State Make sure $s_1$ and $s_2$ are included in $R$ or $NR$
        \If{$s_1, s_2 \in R$}
          \State \Return no
        \Else
          \State \Return \Call{Helper}{$O$, $T$, $I$, $R$, $NR$}
        \EndIf
      \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\newcommand{\IsIconfluent}{\textsc{IsIconfluent}}
\newcommand{\IsIclosed}{\textsc{IsIclosed}}
\newcommand{\Helper}{\textsc{Helper}}

This example motivates \algoref{InteractiveDecisionProcedure}, a general
purpose interactive \Iconfluence{} decision procedure in which a user
iteratively refines the invariant until the decision procedure terminates.
Given an object $O$, a set of transactions $T$, and an invariant $I$, we call
\IsIconfluent$(O, T, I)$ to determine if $T$ is \Iconfluent{}.

\IsIconfluent{} assumes access to an \Iclosed{} decision procedure
\IsIclosed$(O, T, I)$ which returns whether $O$ is \Iclosed{} (yes, no, or
unknown). If it returns no, it also returns a counterexample $s_1$ and $s_2$
such that $I(s_1)$ and $I(s_2)$, but $\lnot I(s_1 \join s_2)$.  Constructing an
\Iclosed{} decision procedure is hard (undecidable, I think), but we can do
pretty well using an SMT solver like Z3. There is also some existing research
on how to built smart \Iclosed{} decision procedures that we can
leverage~\cite{li2014automating}.

\IsIconfluent{} takes in $O$, $T$, and $I$. Formally, $I$ is set of states
(e.g. $\setst{(x, y)}{x \in \ints, y \in \ints, xy \leq 0}$), but an actual
implementation of \IsIconfluent{} would likely take in $I$ as a logical formula
(e.g. $xy \leq 0$). From now on, whenever we mention a (possibly infinite) set
of states, remember that we can always represent the set as a finite formula.

\IsIconfluent{} calls a helper function \Helper{} which---in addition to $O$,
$T$, and $I$---takes as arguments a set $R$ of reachable states and a set $NR$
of unreachable states. Like \IsIconfluent, \Helper$(O, T, I, R, NR)$ returns
whether $T$ is \Iconfluent. Intuitively, $NR$ helps remove unreachable states
from $I$ moving us closer to the case where $I \subseteq
\setst{e}{\reachable{e}}$.

\Helper{} first checks to see if $O$ is \iclosed{$(I - NR)$}. There are three
cases to consider.
\begin{itemize}
  \item
    If \IsIclosed{} gets stuck and doesn't know whether or not $O$ is
    \iclosed{$(I - NR)$}, then \Helper{} is stuck too and is forced to return
    unknown.
  \item
    If \IsIclosed{} determines that $O$ is \iclosed{$(I - NR)$}, then by
    \clmref{StateBasedSufficient}, $T$ is \iconfluent{$(I - NR)$}, so
    $\setst{e}{\reachable{e}} \subseteq I - NR$. $I - NR \subseteq I$, so $T$
    is therefore also \Iconfluent{}. Thus, \Helper{} return yes.
  \item
    If \IsIclosed{} determines that $O$ is not \iclosed{$(I - NR)$}, then we
    have a counterexample $s_1$ and $s_2$. We want to determine whether $s_1$
    and $s_2$ are reachable or unreachable. We can do so in two ways.

    First, we can randomly generate a set of reachable states and add them to
    $R$.

    Second, we can prompt the user to tell us whether or not the states are
    reachable or unreachable. In addition to labelling $s_1$ and $s_2$ as
    reachable or not, the user can also refine $I$ by augmenting $NR$. Taking
    \exampleref{StateBasedNotNecessary} as an example, the user might look
    through the existing reachable and unreachable states and have an aha
    moment realizing that every point with a negative $x$-coordinate is
    unreachable. They can use this knowledge to add all points with negative
    $x$-coordinates to $NR$.

    After $s_1$ and $s_2$ have been labelled reachable or not, we continue. If
    both $s_1$ and $s_2$ are reachable, then so is $s_1 \join s_2$, but $\lnot
    I(s_1 \join s_2)$. Thus, $T$ is not \Iconfluent{}, so \Helper{} returns no.
    Otherwise, one of $s_1$ and $s_2$ is unreachable. Now, we recurse.
\end{itemize}

Let's walk through how we might use \IsIconfluent{} on
\exampleref{StateBasedNotNecessary}.
\begin{itemize}
  \item
    \IsIconfluent{} calls \Helper{} with $R = {(0, 0)}, NR = \emptyset{}$.

  \item
    \Helper{} determines that $O$ is not \iclosed{$(I - R)$} with
    counterexample $s_1 = (-1, 1)$ and $s_2 = (1, -1)$. We take the third
    branch.

  \item
    We randomly generate some reachable points and update $R$ to $\set{(0, 0),
    (1, 0), (0, -1), (1, -1)}$. Lucky for us, $s_2 \in R$, so we know it is
    reachable. We're not sure about $s_1$, so we ask the user. After some
    thought, the user tells us that $s_1$ is unreachable. We recurse with $R =
    \set{(0, 0), (1, 0), (0, -1), (1, -1)}$ and $NR = \set{(-1, 1)}$.

  \item
    \Helper{} determines that $O$ is not \iclosed{$(I - R)$} with
    counterexample $s_1 = (-1, 2)$ and $s_2 = (10, -10)$. We take the third
    branch.

  \item
    We randomly generate some more reachable points and update $R$ to $\set{(0,
    0), (1, 0), (0, -1), (1, -1), (2, 0), (0, -2)}$. $s_1, s_2 \notin R, NR$,
    so we ask the user to label them. The user puts $s_1$ in $NR$ and $s_2$ in
    $R$. We recurse with $R = \set{(0, 0), (1, 0), (0, -1), (1, -1), (2, 0),
    (0, -2), (10, -10)}$ and $NR = \set{(-1, 1), (-1, 2)}$.

  \item
    \Helper{} determines that $O$ is not \iclosed{$(I - R)$} with
    counterexample $s_1 = (-2, 1)$ and $s_2 = (4, -3)$. We take the third
    branch.

  \item
    We randomly generate some more reachable points and update $R$. $s_1, s_2
    \notin R, NR$, so we ask the user to label them. The user notices a pattern
    in $R$ and $NR$ and after some thought, concludes that every point with
    negative $x$-coordinate is unreachable. They update $NR$ to $-\ints \times
    \ints$. We recurse.

  \item
    \Helper{} determines that $O$ is \iclosed{$(I - R)$} and returns yes!
\end{itemize}

### 2018-02-27 doc/paper_redux/new_statebased_sufficient.tex
Show labels.
Finished interactive decision procedure section.

\section{New Sufficient State-Based Conditions}
In this section, we outline a new sufficient condition for state-based
\Iconfluence{} that is more general than the one in
\clmref{statebasedsufficient}. To prove that the condition is sufficient, we
introduce the concepts of \dIconfluence{} and \dIreduction{}.

@@

\begin{claim}\clmlabel{diconfluence_direducible_implies_iconfluence}
  If $T$ is \dIconfluent{} and \dIreducible, then it is \Iconfluent.
\end{claim}
\begin{proof}
  First, we prove that for all $e$, if $\Irec{e}$ then there exists a join-free
  $e'$ where $\Irec{e'}$ and $eval(e) = eval(e')$. We proceed by a pretty
  straightforward structural induction on $e$.
  \begin{itemize}
    \item \textbf{Case 1: $e = s_0$.}
      Trivially, $e' = s_0.$

@@

  To show that $T$ is \Iconfluent{}, it suffices to show that $T$ is
  \dIconfluent{} and \dIreducible{} (by
  \clmref{diconfluence_direducible_implies_iconfluence}). We begin by showing
  that $T$ is \dIconfluent{}. Let
  \[
    e_1 = t_n(t_{n-1}(\ldots (t_1(s_0))\ldots))
    \quad \text{and} \quad
    e_2 = u_m(u_{m-1}(\ldots (u_1(s_0))\ldots))
  \]
  be two arbitrary join-free expressions that recursively satisfy $I$. For ease
  of notation, let
  \[
    \bart{i} = t_i(\ldots(t_1(s_0))\ldots)
    \quad\text{and}\quad
    \baru{i} = u_j(\ldots(u_1(s_0))\ldots)
  \]
  We prove by strong induction on $m \in \nats$ that if $m = i + j$ where $0
  \leq i \leq n$ and $0 \leq j \leq m$, then $\Irec{\bart{i}(\baru{j}(s_0))}$.
  From this, it's immediate that $\Irec{\bart{n}(\baru{m}(s_0))}$, so
  $\Irec{e_1 \join e_2}$, which proves that $T$ is \dIconfluent{}.
  \begin{itemize}
    \item \textbf{Case $m = 0$.}
      $i = j = 0$, so $\bart{0}(\baru{0}(s_0)) = s_0$. $\Irec{s_0}$ because
      $\Irec{e_1}$.

>>>>

\section{New Sufficient State-Based Conditions}
\todo{Motivate these in terms of the interactive decision procedure.}

In this section, we outline a new sufficient condition for state-based
\Iconfluence{} that is more general than the one in
\clmref{StateBasedSufficient}. To prove that the condition is sufficient, we
introduce the concepts of \dIconfluence{} and \dIreduction{}.

@@

\begin{claim}\clmlabel{DiconfluenceDireducibleImpliesIconfluence}
  If $T$ is \dIconfluent{} and \dIreducible, then it is \Iconfluent.
\end{claim}
\begin{proof}
  First, we prove that for all $e$, if $\Irec{e}$ then there exists a join-free
  $e'$ where $\Irec{e'}$ and $eval(e) = eval(e')$. We proceed by a pretty
  straightforward structural induction on $e$.
  \begin{itemize}
    \item \textbf{Case 1: $e = s_0$.}
      Trivially, $e' = s_0.$

@@

  To show that $T$ is \Iconfluent{}, it suffices to show that $T$ is
  \dIconfluent{} and \dIreducible{} (by
  \clmref{DiconfluenceDireducibleImpliesIconfluence}). We begin by showing
  that $T$ is \dIconfluent{}. Let
  \[
    e_1 = t_n(t_{n-1}(\ldots (t_1(s_0))\ldots))
    \quad \text{and} \quad
    e_2 = u_m(u_{m-1}(\ldots (u_1(s_0))\ldots))
  \]
  be two arbitrary join-free expressions that recursively satisfy $I$. For ease
  of notation, let
  \[
    \bart{i} = t_i(\ldots(t_1(s_0))\ldots)
    \quad\text{and}\quad
    \baru{i} = u_j(\ldots(u_1(s_0))\ldots)
  \]
  We prove by strong induction on $m \in \nats$ that if $m = i + j$ where $0
  \leq i \leq n$ and $0 \leq j \leq m$, then $\Irec{\bart{i}(\baru{j}(s_0))}$.
  From this, it's immediate that $\Irec{\bart{n}(\baru{m}(s_0))}$, so
  $\Irec{e_1 \join e_2}$, which proves that $T$ is \dIconfluent{}.
  \begin{itemize}
    \item \textbf{Case $m = 0$.}
      $i = j = 0$, so $\bart{0}(\baru{0}(s_0)) = s_0$. $\Irec{s_0}$ because
      $\Irec{e_1}$.
